
title: "Modelling Extreme Precipitation Events in Southern France"
author: "Raissa YAMEOGO and Flora AUTER"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
  pdf_document:
    toc: true
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5
)

required_packages <- c(
  "data.table",
  "dplyr",
  "extRemes",
  "ggplot2",
  "leaflet",
  "lubridate",
  "POT",
  "RColorBrewer"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  stop(
    "Please install the following packages before running this notebook: ",
    paste(missing_packages, collapse = ", ")
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))
set.seed(123)
```

# 1. Project overview

This project studies historical and projected precipitation data in Southern
France, with a particular focus on extreme rainfall events.

The analysis combines:

- exploratory data analysis;
- spatial visualisation;
- geographic segmentation based on event frequency;
- Gamma generalised linear models;
- cross-validation;
- peaks-over-threshold modelling;
- stationary and non-stationary Generalised Pareto models.

> **Data note.** The source data are not included in this repository.  
> To reproduce the analysis, place the historical and projected datasets in
> the `data/` directory and update the file paths in Section 3.

# 2. Configuration

```{r configuration}
historical_file <- "data/historical_precipitation.txt"
projection_file <- "data/projected_precipitation.txt"

target_latitude <- 43.9999
target_longitude <- 3.08447

precipitation_thresholds <- c(100, 130, 150, 170, 200)
extreme_value_threshold <- 120
```

# 3. Data loading and preparation

## 3.1 Helper functions

```{r helper-functions}
read_historical_data <- function(path) {
  if (!file.exists(path)) {
    stop("Historical data file not found: ", path)
  }

  data <- read.table(path, header = FALSE)

  expected_columns <- c(
    "date",
    "latitude",
    "longitude",
    "precipitation",
    "wind"
  )

  if (ncol(data) != length(expected_columns)) {
    stop(
      "Unexpected number of columns in the historical dataset. ",
      "Expected ", length(expected_columns), ", found ", ncol(data), "."
    )
  }

  names(data) <- expected_columns
  data$date <- as.Date(as.character(data$date), format = "%Y%m%d")

  # Convert precipitation from kg/m²/s to mm/day.
  data$precipitation <- data$precipitation * 86400

  data
}

read_projection_data <- function(path) {
  if (!file.exists(path)) {
    stop("Projection data file not found: ", path)
  }

  data <- read.table(path, header = FALSE)

  expected_columns <- c(
    "date",
    "latitude",
    "longitude",
    "temperature",
    "precipitation",
    "wind"
  )

  if (ncol(data) != length(expected_columns)) {
    stop(
      "Unexpected number of columns in the projection dataset. ",
      "Expected ", length(expected_columns), ", found ", ncol(data), "."
    )
  }

  names(data) <- expected_columns
  data$date <- as.Date(as.character(data$date), format = "%Y%m%d")

  data
}

add_calendar_variables <- function(data) {
  data %>%
    mutate(
      year = year(date),
      month = factor(month(date)),
      day = day(date),
      year_month = format(date, "%Y-%m")
    )
}
```

## 3.2 Load the datasets

```{r load-data}
historical_data <- read_historical_data(historical_file)
projection_data <- read_projection_data(projection_file)

# Keep only variables shared by both datasets before concatenation.
projection_for_merge <- projection_data %>%
  select(date, latitude, longitude, precipitation, wind)

weather_data <- bind_rows(
  historical_data,
  projection_for_merge
) %>%
  add_calendar_variables()

glimpse(weather_data)
```

# 4. Analysis of a selected geographic point

```{r selected-location}
location_data <- weather_data %>%
  filter(
    latitude == target_latitude,
    longitude == target_longitude
  )

if (nrow(location_data) == 0) {
  stop(
    "No observation was found for the selected coordinates: ",
    target_latitude, ", ", target_longitude
  )
}
```

## 4.1 Location map

```{r location-map}
leaflet() %>%
  addTiles() %>%
  setView(
    lng = target_longitude,
    lat = target_latitude,
    zoom = 12
  ) %>%
  addCircleMarkers(
    lng = target_longitude,
    lat = target_latitude,
    radius = 8,
    weight = 2,
    opacity = 0.8,
    popup = paste(
      "Latitude:", target_latitude,
      "<br>Longitude:", target_longitude
    )
  )
```

## 4.2 Precipitation over time

```{r precipitation-time-series}
ggplot(location_data, aes(x = date, y = precipitation)) +
  geom_line(alpha = 0.6) +
  labs(
    title = "Daily precipitation over time",
    x = "Date",
    y = "Precipitation (mm/day)"
  ) +
  theme_minimal()
```

## 4.3 Distribution of precipitation

```{r precipitation-distribution}
ggplot(location_data, aes(x = precipitation)) +
  geom_histogram(bins = 60) +
  labs(
    title = "Distribution of daily precipitation",
    x = "Precipitation (mm/day)",
    y = "Number of observations"
  ) +
  theme_minimal()
```

```{r precipitation-distribution-above-20}
location_data %>%
  filter(precipitation >= 20) %>%
  ggplot(aes(x = precipitation)) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of precipitation above 20 mm/day",
    x = "Precipitation (mm/day)",
    y = "Number of observations"
  ) +
  theme_minimal()
```

## 4.4 Number of threshold exceedances

```{r exceedance-counts}
threshold_sequence <- seq(80, 500, by = 10)

exceedance_counts <- data.frame(
  threshold = threshold_sequence,
  number_of_exceedances = vapply(
    threshold_sequence,
    function(threshold) {
      sum(location_data$precipitation > threshold, na.rm = TRUE)
    },
    numeric(1)
  )
) %>%
  filter(number_of_exceedances > 0)

exceedance_counts
```

## 4.5 Annual and monthly summaries

```{r annual-summary}
annual_summary <- location_data %>%
  group_by(year) %>%
  summarise(
    mean_precipitation = mean(precipitation, na.rm = TRUE),
    maximum_precipitation = max(precipitation, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(annual_summary, aes(x = year, y = maximum_precipitation)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Annual maximum precipitation",
    x = "Year",
    y = "Maximum precipitation (mm/day)"
  ) +
  theme_minimal()
```

```{r monthly-summary}
monthly_summary <- location_data %>%
  group_by(year_month) %>%
  summarise(
    mean_precipitation = mean(precipitation, na.rm = TRUE),
    maximum_precipitation = max(precipitation, na.rm = TRUE),
    .groups = "drop"
  )

head(monthly_summary)
```

# 5. Spatial frequency of extreme precipitation

## 5.1 Frequency calculation

```{r spatial-frequency-function}
compute_annual_exceedance_frequency <- function(data, threshold) {
  data %>%
    mutate(exceedance = as.integer(precipitation > threshold)) %>%
    group_by(latitude, longitude, year) %>%
    summarise(
      annual_exceedances = sum(exceedance, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    group_by(latitude, longitude) %>%
    summarise(
      mean_annual_exceedances = mean(annual_exceedances, na.rm = TRUE),
      .groups = "drop"
    )
}
```

```{r spatial-frequency}
spatial_frequency <- compute_annual_exceedance_frequency(
  weather_data,
  threshold = extreme_value_threshold
)

head(spatial_frequency)
```

## 5.2 Geographic segmentation

The points are grouped according to their average annual number of threshold
exceedances.

```{r geographic-segmentation}
spatial_clusters <- spatial_frequency %>%
  mutate(
    cluster = case_when(
      mean_annual_exceedances > 1 ~ "High frequency",
      mean_annual_exceedances > 0.5 ~ "Moderate frequency",
      mean_annual_exceedances > 0.1 ~ "Low frequency",
      TRUE ~ "Very low frequency"
    ),
    cluster = factor(
      cluster,
      levels = c(
        "High frequency",
        "Moderate frequency",
        "Low frequency",
        "Very low frequency"
      )
    )
  )

table(spatial_clusters$cluster)
```

```{r cluster-map}
cluster_palette <- colorFactor(
  palette = "RdYlGn",
  domain = spatial_clusters$cluster,
  reverse = TRUE
)

leaflet(spatial_clusters) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = ~pmax(3, 2 + mean_annual_exceedances),
    color = ~cluster_palette(cluster),
    stroke = FALSE,
    fillOpacity = 0.7,
    popup = ~paste0(
      "Cluster: ", cluster,
      "<br>Mean annual exceedances: ",
      round(mean_annual_exceedances, 2)
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = cluster_palette,
    values = ~cluster,
    title = "Extreme-event frequency"
  )
```

# 6. Gamma GLM modelling

The objective is to explain strictly positive precipitation using available
meteorological and calendar covariates.

## 6.1 Cross-validation function

```{r glm-cross-validation}
cross_validate_gamma_glm <- function(formula, data, k = 10, seed = 123) {
  model_data <- data %>%
    filter(precipitation > 0) %>%
    tidyr::drop_na(all.vars(formula))

  if (nrow(model_data) < k) {
    stop("The number of observations must be greater than k.")
  }

  set.seed(seed)
  fold_id <- sample(rep(seq_len(k), length.out = nrow(model_data)))

  rmse <- numeric(k)

  for (fold in seq_len(k)) {
    training_data <- model_data[fold_id != fold, , drop = FALSE]
    validation_data <- model_data[fold_id == fold, , drop = FALSE]

    fitted_model <- glm(
      formula = formula,
      data = training_data,
      family = Gamma(link = "log")
    )

    predictions <- predict(
      fitted_model,
      newdata = validation_data,
      type = "response"
    )

    rmse[fold] <- sqrt(
      mean(
        (predictions - validation_data$precipitation)^2,
        na.rm = TRUE
      )
    )
  }

  full_model <- glm(
    formula = formula,
    data = model_data,
    family = Gamma(link = "log")
  )

  data.frame(
    formula = paste(deparse(formula), collapse = ""),
    AIC = AIC(full_model),
    BIC = BIC(full_model),
    RMSE = mean(rmse)
  )
}
```

## 6.2 Candidate models

```{r candidate-models}
candidate_formulas <- list(
  precipitation ~ wind,
  precipitation ~ month,
  precipitation ~ year,
  precipitation ~ wind + month,
  precipitation ~ wind + year,
  precipitation ~ month + year,
  precipitation ~ wind + month + year
)

glm_results <- bind_rows(
  lapply(
    candidate_formulas,
    cross_validate_gamma_glm,
    data = location_data,
    k = 10
  )
) %>%
  arrange(RMSE)

glm_results
```

```{r best-glm}
best_formula <- candidate_formulas[[
  which.min(vapply(
    candidate_formulas,
    function(formula) {
      cross_validate_gamma_glm(
        formula = formula,
        data = location_data,
        k = 10
      )$RMSE
    },
    numeric(1)
  ))
]]

best_gamma_glm <- glm(
  formula = best_formula,
  data = location_data %>% filter(precipitation > 0),
  family = Gamma(link = "log")
)

summary(best_gamma_glm)
```

# 7. Extreme value analysis

## 7.1 Threshold diagnostics

```{r threshold-diagnostics}
positive_precipitation <- location_data$precipitation[
  location_data$precipitation > 0
]

mrlplot(
  positive_precipitation,
  nt = 50,
  tlim = c(80, 130)
)

tcplot(
  positive_precipitation,
  tlim = c(80, 130),
  nt = 50
)
```

The final threshold should be selected using both graphical diagnostics and
the number of observations remaining above the threshold.

## 7.2 Stationary GPD model

```{r stationary-gpd}
stationary_gpd <- fitgpd(
  positive_precipitation,
  threshold = extreme_value_threshold
)

plot(stationary_gpd)
stationary_gpd
```

## 7.3 Non-stationary GPD model

The following model allows the scale parameter to vary with time.

```{r non-stationary-gpd}
extreme_value_data <- location_data %>%
  mutate(year_numeric = as.numeric(year))

non_stationary_gpd <- fevd(
  x = precipitation,
  data = extreme_value_data,
  threshold = extreme_value_threshold,
  scale.fun = ~ year_numeric,
  type = "GP",
  use.phi = TRUE,
  verbose = TRUE
)

plot(non_stationary_gpd)
non_stationary_gpd
```

# 8. Model comparison and interpretation

For model selection:

- use RMSE to assess out-of-sample predictive performance;
- use AIC and BIC to compare model fit while penalising complexity;
- inspect diagnostic plots for Gamma GLMs and GPD models;
- verify that parameter estimates remain stable across reasonable thresholds;
- avoid drawing causal conclusions from statistical associations alone.

# 9. Main limitations

This analysis has several limitations:

1. the quality of the conclusions depends on the climate model and the spatial
   resolution of the data;
2. nearby geographic points may exhibit spatial dependence;
3. the fixed threshold used for the GPD analysis may not be optimal for every
   location;
4. non-stationarity may require richer time-varying structures;
5. uncertainty from climate projections should be propagated into the final
   risk estimates.

# 10. Reproducibility

To make the project fully reproducible:

```r
install.packages("renv")
renv::init()
renv::snapshot()
```

The resulting `renv.lock` file should be committed to the repository.
